Fused regression for network inference in multiple species

### 2015-08-27 paper1.tex
changed intro

\section{Introduction}
As the volume and variety of genome scale data have exploded, the goal of accurately modeling gene regulatory networks has become more attainable. Already, there is enough data in some model organisms for the development of high quality networks; however, this is still not true of many other model systems. Luckily, the problems of inferring gene regulatory networks in multiple species are not independent, due to the presence of conserved structure. There has been a great deal of work showing that functional conservation exists in gene regulatory networks even across large evolutionary distance. Abundant data in one species should be a useful source when learning the network of a related species. But most existing inference algorithms do not incorporate this information.

We employ a linear model of gene expression in which the rate of expression is assumed to be proportional to a weighted sum of transcription factor expression. These weights represent the regulatory interactions we wish to learn. We can fit these weights simultaneously in multiple species of interest using time series and steady state expression data. This is similar to many existing network inference methods and produces identical answers to fitting each species separately. To take advantage of cross-species information, we introduce additional constraints on the problem that penalize differences between regulatory weights which are expected to be similar. The choice of which interactions to fuse together follows from the observation that orthologous genes are likely to have similar regulatory interactions.

A few approaches to multi-source network inference have been proposed; however, current approaches either require that the sources have identical sets of genes [aviv regev paper], which is useful for analyzing cell lines but not for most cross species comparisons, or restrict analysis to the subset of genes with orthologs in every species [ibm paper], which limits the applicability of the model. Our approach allows simultaneous inference of the whole gene regulatory networks of species that may have partial orthology overlap. We demonstrate performance gains in networks learned from simulated data and real data from Bacillus subtilis and Bacillus anthracis, where a large fraction of genes have no orthologs. Though this method was developed for multi-species network inference, it is also useful in the easier problem of learning networks in multiple cell lines.

The largest gains in performance are obtained when there is very little data in a species of interest but a large amount of data in a related species. But the information flow is two ways and our simulation results suggest that network inference improves in both species. This feature of simultaneously fitting networks is particularly useful when there is an adequate amount of data available but spread over too many species.

There exist several tools for identifying orthologous genes using sequence similarity. Although sequence similarity is a useful predictor of functional similarity, we recognize that some genes have taken on different functions and therefore may have new regulatory interactions. When this occurs, the benefits in overall performance from our method come at the cost of impaired prediction of regulators for these genes. The constraints we impose favor weight configurations in which fused interaction weights are similar to one another. Our method minimizes a cost function that strives  to simultaneously fit expression data in each species and produce networks that are consistent with these evolutionary constraints. When a gene's regulators have changed through evolution, there may be no way of achieving both these goals. In this case, the network inference procedure would be improved by relaxing or ignoring these evolutionary constraints. However, we don't know a priori which constraints to enforce and which to relax.

@@

We present two methods to multi-species network inference using fused regression. One approach, our fused L2 method, uses an L2 fusion penalty on differences in regulatory weights with evolutionary constraints. Additionally, we developed a method, which we call adaptive fusion, using a non-convex saturating fusion penalty for both network inference and network comparison. We test the ability of fused L2 and adaptive fusion to improve network recovery on both synthetic data and Bacillus subtilis and Bacillus anthracis. We explore the circumstances under which each approach is optimal, and evaluate the robustness of adaptive fusion to incorrect orthology, simulating the biologically relevant case of neofunctionalization.

>>>>

\section{Introduction}
As the volume and variety of genome scale data have exploded, the goal of accurately modeling gene regulatory networks has become more attainable. Large-scale data collection efforts have contributed to the development of high quality networks, but there remain many biological processes and organisms for which there does not exist enough data. Furthermore, as new technologies are developed and some old ones are replaced, such as RNAseq and microarray, it becomes important to be able to combine different data sets, lest we lose valuable information from previous studies. The problems of inferring networks in related species, processes, and using similar technologies likely are not independent ones, and benefit from the infusion of additional data.

We present two methods for network inference which use fused regression, a type of penalized least squares regression, in which the difference of regulatory weights which are thought to be similar are penalized. Our fused L2 method uses a ridge fusion penalty on these differences, and is useful where the relationships between nodes (similarity of genes between data sets) is reliable. Our adaptive fusion uses a non-concave saturating fusion penalty to simultaneously learn multiple networks and conservation of these networks.

To learn gene regulatory networks, we employ a linear model of gene expression in which the rate of expression is assumed to be proportional to a weighted sum of transcription factor expression. These weights represent the regulatory interactions we wish to learn. We can fit these weights simultaneously in multiple data sources of interest using time series and steady state expression data. This is similar to many existing network inference methods and produces identical answers to fitting each network separately. To take advantage of cross-dataset information, we introduce additional constraints on the problem that penalize differences between regulatory weights which are expected to be similar. Genes are not always regulated by the same transcription factors under different conditions. For example, a TF which promotes expression of a gene under a given process may repress that same gene at another point. Using data from these conditions to learn one network will prevent the identification of these contradictory regulatory interactions; it is therefore important to be able to combine data in a principled manner. We can use knowledge about the conservation of regulation to while leveraging th eadditional data to learn networks describing disparate processes. In the case of multiple species, there has been a great deal of work showing that functional conservation exists in gene regulatory networks even across large evolutionary distance. We make the assumption that if a transcription factor in one species is orthologous to a TF in a related species, and a gene in the first species is orthologous to a gene in the second, then the interaction between the TF and gene in the first species should be similar to the interaction between the orthologous TF, gene pair in the second species.

There exist several tools for identifying orthologous genes using sequence similarity. Although sequence similarity is a useful predictor of functional similarity, we recognize that some genes have taken on different functions and therefore may have new regulatory interactions. When this occurs, the benefits in overall performance from our L2 fusion approach come at the cost of impaired prediction of regulators for these genes. The constraints we impose favor weight configurations in which fused interaction weights are similar to one another. Our fused L2 method minimizes a cost function that strives to simultaneously fit expression data in each species and produce networks that are consistent with these evolutionary constraints. When a gene's regulators have changed through evolution, there may be no way of achieving both these goals. In this case, the network inference procedure would be improved by relaxing or ignoring these evolutionary constraints. However, we don't know a priori which constraints to enforce and which to relax.

@@

We test the ability of fused L2 and adaptive fusion to improve network recovery on both synthetic data and Bacillus subtilis and Bacillus anthracis. We explore the circumstances under which each approach is optimal, and evaluate the robustness of adaptive fusion to incorrect orthology, simulating the biologically relevant case of neofunctionalization.
